---
title: "Examining the Impact of Environmental and Demographic Factors on Marathon Performance: An Exploratory Data Analysis"
author: Michael Stewart
date: October 6, 2024
format:
  pdf:
    documentclass: article
    geometry: margin= .5in
editor: visual
---

# Abstract

This study investigates the impact of age, gender, and environmental conditions on marathon performance using data from 11,564 athletes aged 14 to 85 across five major races: Boston, Chicago, New York, Twin Cities, and Duluth (Grandma’s). The analysis aimed to explore three key objectives: (1) the effects of aging on marathon performance in men and women, (2) the influence of environmental factors, such as weather and air quality, on performance, and (3) the identification of the most impactful environmental factors.

Results demonstrate that marathon performance peaks for both men and women in the 26-35 age range, with performance declining progressively in older age groups. Gender differences were evident, with men consistently outperforming women across all age bins. This gender gap became more pronounced in older groups, especially in the 65+ age category. Notably, in Duluth’s summer marathon, women appeared to reach peak performance six years later than men, suggesting potential gendered differences in thermoregulation and heat adaptation.

Environmental factors, particularly WBGT and race-day flag conditions, significantly influenced marathon performance. Red flag conditions, representing the most severe weather, were associated with the largest performance declines, especially among older runners. By contrast, air quality, measured as ozone concentration, showed no significant relationship with performance, likely due to relatively low levels during race events. While WBGT and dry bulb temperature were the most impactful continuous variables, their effects were modest, highlighting the importance of categorical measures like flag conditions for assessing weather-related impacts.

# Introduction

This project, conducted in collaboration with Dr. Brett Romano Ely and Dr. Matthew Ely from Providence College, examines the effects of environmental conditions on marathon performance across age and gender. The dataset includes performances from five major marathons (Boston, New York, Chicago, Twin Cities, Grandma’s) with detailed environmental data such as temperature, humidity, WBGT (wet bulb globe temperature), and air quality, for 11,564 athletes aged 14 to 85.

Marathon performance is influenced by a complex interplay of physiological, demographic, and environmental factors. Previous research has demonstrated that marathon performance declines with age, with peak performance typically occurring in the late 20s to early 30s for both men and women (Ely et al., 2010). However, the rate and extent of decline differ between genders, potentially due to physiological differences such as muscle mass, endurance capacity, and thermoregulatory efficiency (Besson et al., 2022; Yanovich et al., 2020). This study extends these findings by investigating how aging impacts men and women differently, particularly under varying environmental conditions.

Environmental stressors such as heat, humidity, and solar radiation can significantly affect endurance performance, with WBGT emerging as a critical measure of heat stress (Kenney & Munce, 2003). For example, races held in summer conditions, such as Duluth (Grandma’s Marathon), often pose unique challenges due to higher temperatures and humidity levels. While men generally outperform women in endurance events, women’s thermoregulatory advantages under heat stress may allow them to sustain performance under moderate conditions, as observed in other studies (Ely et al., 2007). This study also incorporates air quality measures, specifically ozone concentration, as prior research has suggested potential respiratory limitations associated with air pollution during endurance activities.

Through exploratory data analysis, we investigated three main objectives. First, we assessed the effects of aging on marathon performance in men and women. Second, we explored how environmental conditions, including temperature, humidity, and air quality, impact performance. Third, we identified which environmental factors, such as WBGT and temperature, exert the greatest influence on performance outcomes. Statistical analyses, including the Kruskal-Wallis test and pairwise Wilcoxon comparisons, were employed to evaluate these relationships. These non-parametric methods were selected to account for the non-normality of the data and unequal variances across groups. This approach provides robust insights into the interaction between age, gender, and environmental conditions in shaping endurance performance outcomes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      echo = FALSE,
                      fig.align = "center")
# load libraries
library(tidyverse)
library(gt)
library(lubridate)
library(reshape2)
library(broom)


################## DATA IMPORT AND INITIAL CLEANING ################## 

# read in datasets
course_record <- read.csv("~/Downloads/course_record-2.csv")
marathon_dates <- read.csv("~/Downloads/marathon_dates.csv")
aqi_vals <- read.csv("~/Downloads/aqi_values.csv")
project1 <- read.csv("~/Downloads/project1.csv")

# convert course record times to seconds
course_record <- course_record %>%
  mutate(CR_sec = as.numeric(hms(course_record$CR)))

# map race and sex
race_map <- c("0" = "B", "1" = "C", "2" = "NY", "3" = "TC", "4" = "D")
sex_map <- c("0" = "F", "1" = "M")

# adjust race mapping for duluth (grandma's marathon)
race_map.1 <- c("B" = "Boston", "C" = "Chicago", "NY" = "New York", 
                "TC" = "Twin Cities", "D" = "Duluth (Grandma's)")


################## DATA MERGING ################## 
# apply race and sex mappings to project1
project1 <- project1 %>%
  mutate(
    Race_standard = race_map[as.character(Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D.)],
    Sex_standard = sex_map[as.character(Sex..0.F..1.M.)]
  )



# join project1 with course_record
merged_df <- project1 %>%
  left_join(course_record, by = c("Race_standard" = "Race", "Year", 
                                  "Sex_standard" = "Gender"))

# remove unnecessary columns, calculate time in minutes
merged_df <- merged_df %>%
  select(-Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D., -Sex..0.F..1.M.) %>%
  mutate(time_min = (CR_sec * (1 + (X.CR / 100))) / 60)

# rename columns for clarity
merged_df <- merged_df %>%
  rename(
    race = Race_standard,
    year = Year,
    sex = Sex_standard,
    flag = Flag,
    age = Age..yr.,
    percent_cr = X.CR,
    dew_point = DP,
    rel_humid = X.rh,
    globe_temp = Tg..C,
    wb_temp = Tw..C,
    db_temp = Td..C,
    solar_rad = SR.W.m2,
    wind = Wind
  ) %>%
  mutate(race = race_map.1[race])

merged_df <- merged_df %>%
  mutate(age_bin = cut(age, 
                       breaks = c(0, 18, 26, 36, 46, 56, 65, Inf),
                       labels = c("< 18", "18-25", "26-35", "36-45", 
                                  "46-55", "56-64", "65+"),
                       right = FALSE))

# ensure race names are consistent between marathon_dates and merged_df
marathon_dates <- marathon_dates %>%
  mutate(race = case_when(
    marathon == "NYC" ~ "New York",               
    marathon == "Grandmas" ~ "Duluth (Grandma's)", 
    T ~ marathon  
  ))

# join with marathon_dates
marathon_dates <- marathon_dates %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"))

merged_df <- merged_df %>%
  left_join(marathon_dates, by = c("race", "year"))


######### MERGE AQI VAL INFO ##################

# standardize race names in aqi_vals
aqi_vals <- aqi_vals %>%
  rename(race = marathon) %>%
  mutate(
    race = case_when(
      race == "NYC" ~ "New York",              
      race == "Grandmas" ~ "Duluth (Grandma's)",  
      T ~ race  
    ),
    date = as.Date(date_local, format = "%Y-%m-%d"),
    year = as.numeric(format(date, "%Y"))
  ) %>%
  select(-date_local)

# calculate average ozone ppm (8-hour avg)
avg_ppm <- aqi_vals %>%
  filter(units_of_measure == "Parts per million", 
         sample_duration == "8-HR RUN AVG BEGIN HOUR") %>%
  group_by(race, year, date) %>%
  summarize(avg_ppm = mean(arithmetic_mean, na.rm = T)) %>%
  ungroup()

# join ppm data with merged_df, final join
merged_df <- merged_df %>%
  left_join(avg_ppm, by = c("race", "year", "date"))

```

# Methods

## Missing Data Handling

Missing weather data were addressed through a complete case analysis, removing rows with missing values for key environmental variables, such as wet bulb temperature, dry bulb temperature, relative humidity, and air quality (ozone PPM). The proportion of missing data was minimal, accounting for only 4.25% of total observations, as shown in **Table 0**. For instance, in 2011, the Chicago marathon had 126 missing observations (1.09%), New York had 131 (1.13%), and Twin Cities had 118 (1.02%). Duluth (Grandma’s) marathon in 2012 had 116 missing values (1.00%). Given the relatively small percentage of missing data, this approach ensured the integrity of the data set while avoiding unnecessary complexity in imputation. Dropping these missing values yielded 11,073 participants left to be analyzed.

```{r}

########## EXPLORING MISSING DATA ################## 

# weather related variables
weather_vars <- c("db_temp", "wb_temp", "rel_humid", "globe_temp", "solar_rad", 
                  "dew_point", "wind", "WBGT")

# total observations in the data
total_observations <- nrow(merged_df)

# filter missing weather data
missing_weather <- merged_df %>%
  filter_at(vars(weather_vars), any_vars(is.na(.))) %>%
  group_by(year, race) %>%
  summarise(
    missing_count = n()  # number ofmissing rows for each year and race
  ) %>%
  mutate(missing_percentage = (missing_count / total_observations) * 100) %>%
  arrange(year, race)

# total missing data for all years and races combined
total_missing <- sum(missing_weather$missing_count)
total_missing_percentage <- (total_missing / total_observations) * 100

# convert year to character 
missing_weather <- missing_weather %>%
  mutate(year = as.character(year))  

# total summary row at the bottom of the table
missing_weather <- missing_weather %>%
  bind_rows(
    tibble(
      year = "Total",
      race = "",
      missing_count = total_missing,
      missing_percentage = total_missing_percentage
    )
  )

# table with percentage of missing data and bold years
missing_weather_table <- missing_weather %>%
  gt() %>%
  tab_header(
    title = "Table 0: Summary of Missing Weather Data by Year and Race",
    subtitle = "Missing Data Relative to Total Observations"
  ) %>%
  cols_label(
    year = "Year",
    race = "Race",
    missing_count = "Missing Data Count",
    missing_percentage = "Missing Data (%)"
  ) %>%
  fmt_number(
    columns = vars(missing_count),
    decimals = 0
  ) %>%
  fmt_number(
    columns = vars(missing_percentage),
    decimals = 2
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      columns = vars(year)
    )
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = year == "Total"
    )
  )

missing_weather_table

```

## Data Preparation

The dataset was comprised of marathon performance data from five major races, along with corresponding environmental conditions, including air quality measures. Marathon performance data were merged with course records, race dates, and air quality data. The primary air quality variable, ozone concentration, was measured in parts per million (PPM) and calculated as an 8-hour average for each race-year. We chose PPM over micrograms per cubic meter because PPM is the standard unit for measuring gas concentrations like ozone, whereas micrograms per cubic meter is typically used for particulate matter such as PM2.5. The 8-hour average was selected to better reflect longer-term exposure during marathon events, as opposed to shorter-term (e.g., 1-hour) measurements, which may not capture the full extent of air quality conditions experienced by runners throughout the race. This approach provided a standardized metric for comparing air quality across different races and years.

To ensure comparability across different analyses, marathon times were treated in two distinct ways depending on the context. When analyses did not group or facet by race, marathon times were represented in minutes, allowing for more straightforward and interpretable results. However, when grouping or comparing across races, percent course record was used (percent CR) to account for differences in race conditions and course difficulty. Percent CR standardizes a runner’s time as a percentage of the course record for each marathon. A value of 0% means the runner matched the course record, while positive values indicate slower times, and negative values represent faster times. This allows performance to be compared across different races and conditions.

```{r}
###### DATA CLEANING; DROP NAs ################## 
cleaned_df <- merged_df %>%
  drop_na(db_temp, wb_temp, rel_humid, globe_temp, 
          solar_rad, dew_point, wind, WBGT)

```

## Exploratory Data Analysis

Exploratory data analysis (EDA) focused on the relationship between marathon performance and the variables of interest: age, gender, and environmental conditions. Marathon times were analyzed using summary statistics and visualizations, such as boxplots, to explore trends across different age groups and genders. This analysis provided insights into how performance varied with age, with comparisons made across men and women.

## Testing Normality

The Shapiro-Wilk test results demonstrate that neither the raw data (percent_cr) nor the square root-transformed data (percent_cr_sqrt) meet the assumption of normality, with p-values consistently below 0.05 across all age bins. This indicates strong evidence of non-normality in the dataset, even after applying a square root transformation. The Q-Q plots further confirm this finding, as both the original and transformed data show clear deviations from the theoretical normal distribution. While the square root transformation mitigates some skewness, it does not adequately normalize the data.

```{r}

cleaned_df <- cleaned_df %>%
  mutate(percent_cr_sqrt = sqrt(percent_cr))

library(ggplot2)
library(dplyr)
library(car)

# Density Plot
density_plot <- ggplot(cleaned_df) +
  geom_density(aes(x = percent_cr, fill = flag), alpha = 0.4) +
  geom_density(aes(x = percent_cr_sqrt, fill = flag), alpha = 0.4, linetype = "dashed") +
  facet_wrap(~race, scales = "free") +
  theme_classic() +
  labs(
    title = "Density Plot of Marathon Performance by Age Bin and Race",
    subtitle = "Solid: Original Scale, Dashed: Square Root Transformed",
    x = "Marathon Performance (% Course Record)",
    y = "Density"
  ) +
  theme(legend.position = "bottom")

print(density_plot)

# Q-Q Plot
qq_plot <- ggplot(cleaned_df, aes(sample = percent_cr)) +
  stat_qq(aes(color = "Original Scale")) +
  stat_qq_line(aes(color = "Original Scale")) +
  stat_qq(aes(sample = percent_cr_sqrt, color = "Square Root Transformed")) +
  stat_qq_line(aes(sample = percent_cr_sqrt, color = "Square Root Transformed")) +
  facet_wrap(~race) +
  theme_classic() +
  labs(
    title = "Q-Q Plots of Marathon Performance by Race",
    subtitle = "Original Scale vs. Square Root Transformed",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles",
    color = "Data Transformation"
  )

print(qq_plot)
```

Levene’s test results for homogeneity of variance reveal that the raw data meets this assumption, with a p-value of 0.618, indicating no significant differences in variance across groups. However, the square root-transformed data fails the test for homogeneity, with a p-value of 0.041, suggesting unequal variances among groups

he density plots reinforce these conclusions, illustrating skewed distributions and potential outliers for both the original and square root-transformed data across races. Taken together, these results indicate that the assumptions required for parametric methods, such as ANOVA, are not satisfied. As a result, we will employ non-parametric methods, specifically the Kruskal-Wallis test and pairwise Wilcoxon tests, to assess group differences. These methods do not require normality or equal variances, making them more appropriate for the characteristics of this dataset.

```{r}

# Levene's Test
levene_raw <- car::leveneTest(percent_cr ~ flag, data = cleaned_df)
levene_sqrt <- car::leveneTest(percent_cr_sqrt ~ flag, data = cleaned_df)

levene_results <- data.frame(
  Test = c("Raw Data", "Square Root Transformed"),
  F_statistic = c(levene_raw$`F value`[1], levene_sqrt$`F value`[1]),
  P_value = c(levene_raw$`Pr(>F)`[1], levene_sqrt$`Pr(>F)`[1])
)

# 1. Shapiro-Wilk Test Results Table
shapiro_results <- cleaned_df %>%
  group_by(age_bin) %>%
  summarize(
    Raw_p = shapiro.test(percent_cr)$p.value,
    Sqrt_p = shapiro.test(percent_cr_sqrt)$p.value
  ) %>%
  mutate(
    Normality_Raw = ifelse(Raw_p < 0.05, "No", "Yes"),
    Normality_Sqrt = ifelse(Sqrt_p < 0.05, "No", "Yes")
  )

shapiro_table <- shapiro_results %>%
  gt() %>%
  tab_header(
    title = "Shapiro-Wilk Test Results for Normality",
    subtitle = "Raw and Square Root Transformed Data"
  ) %>%
  cols_label(
    age_bin = "Age Bin",
    Raw_p = "Raw p-value",
    Sqrt_p = "Sqrt p-value",
    Normality_Raw = "Normality (Raw)",
    Normality_Sqrt = "Normality (Sqrt)"
  ) %>%
  fmt_number(columns = c(Raw_p, Sqrt_p), decimals = 3)

shapiro_table



levene_results <- data.frame(
  Test = c("Raw Data", "Square Root Transformed"),
  F_statistic = c(levene_raw$`F value`[1], levene_sqrt$`F value`[1]),
  P_value = c(levene_raw$`Pr(>F)`[1], levene_sqrt$`Pr(>F)`[1]),
  Homogeneity = ifelse(c(levene_raw$`Pr(>F)`[1], levene_sqrt$`Pr(>F)`[1]) < 0.05, "No", "Yes")
)

levene_table <- levene_results %>%
  gt() %>%
  tab_header(
    title = "Levene's Test Results for Homogeneity of Variance",
    subtitle = "Raw and Square Root Transformed Data"
  ) %>%
  cols_label(
    Test = "Data Type",
    F_statistic = "F-Statistic",
    P_value = "p-value",
    Homogeneity = "Homogeneity of Variance"
  ) %>%
  fmt_number(columns = c(F_statistic, P_value), decimals = 3)

levene_table
```

# Statistical Analysis

To assess the effects of environmental variables and demographic factors on marathon performance, non-parametric methods were employed due to the non-normal distribution of the data and unequal variances across groups. Specifically, the Kruskal-Wallis test was used to examine the influence of gender, age, and race-day flag conditions on performance times. Pairwise Wilcoxon tests, with Bonferroni adjustment, were performed to identify significant differences between groups. These methods allowed for robust comparisons across demographic and environmental factors.

# Results

## Aim 1: Effects of Increasing Age on Marathon Performance in Men and Women

The analysis of marathon performance across age groups and genders reveals clear trends regarding how marathon times are affected by aging. As shown in **Figure 1.1**, marathon times are fastest among runners in the 26-35 age range across all five major marathons. Both men and women exhibit slower times as they age, with performance declining progressively in the older age groups. The gender differences are evident, with men generally running faster than women at all ages. Notably, the gap between male and female performance becomes more pronounced in the older age categories, particularly among runners aged 56-64 and 65+, where women tend to take significantly longer to complete the marathon compared to men. Variability in times also increases as participants age for both sexes. It's important to note that New York and Boston did not record any times for participants in the \<18 age group, which may influence the comparisons for this category across marathons.

```{r fig.width=10, fig.height=7}

age_time_bp <- ggplot(cleaned_df, aes(x = age_bin, y = time_min, fill = sex)) +
  facet_wrap(~race) +
  geom_boxplot(outlier.size = 0.5, size = 0.35) +
  labs(x = "Age", y = "Marathon Time in Minutes",
       title = "Figure 1.1: Marathon Times by Age and Gender",
       fill = "sex") +
  scale_fill_manual(values = c("F" = "#BE398D", "M" = "#43A5BE")) +
  theme_classic(base_family = "serif") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")


age_time_bp

```

**Table 1.1**, summarizing marathon performance by age and sex, further illustrates these trends. For instance, among 26-35 year olds, male marathoners averaged 139.4 minutes (SD = 10.1), while female marathoners averaged 162.8 minutes (SD = 15.4). This pattern persists across all age groups, and by the 65+ category, men were completing marathons in an average of 276 minutes (SD = 67.6) compared to women’s 318.1 minutes (SD = 63.3). This nearly 42-minute difference shows the increasing performance gap as athletes age.

```{r}
marathon_summary <- cleaned_df %>%
  group_by(age_bin) %>%
  summarize(
    # male count and percentage
    male_counts = paste0(n_male <- sum(sex == "M", na.rm = T), 
                         " (", round(mean(sex == "M", na.rm = T) * 100, 1), "%)"),
    
    # female count and percentage
    female_counts = paste0(n_female <- sum(sex == "F", na.rm = T), 
                           " (", round(mean(sex == "F", na.rm = T) * 100, 1), "%)"),
    
    # mean and sd for male times
    male_time = paste0(round(mean(time_min[sex == "M"], na.rm = T), 1), 
                       " (", round(sd(time_min[sex == "M"], na.rm = T), 1), ")"),
    
    # mean and sd for female times
    female_time = paste0(round(mean(time_min[sex == "F"], na.rm = T), 1), 
                         " (", round(sd(time_min[sex == "F"], na.rm = T), 1), ")"),
    
    # total number of participants in each age group
    n_total = n()
  ) %>%
  ungroup()


marathon_summ_table <- marathon_summary %>%
  gt() %>%
  tab_header(
    title = "Table 1.1: Marathon Performance Summary by Age Group and Sex"
  ) %>%
  cols_label(
    age_bin = "Age Group",
    male_counts = "Male (n, %)",
    female_counts = "Female (n, %)",
    male_time = "Male Marathon Time",
    female_time = "Female Marathon Time",
    n_total = "n"
  ) %>%
  cols_align(
    align = "center", 
    columns = everything()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  fmt_number(
    columns = c("n_total"),
    decimals = 0
  ) %>%
  tab_footnote(
    footnote = "Marathon time is presented as mean (SD) in minutes.",
    locations = cells_column_labels(columns = c("male_time", "female_time"))
  )

marathon_summ_table

```

The results of the Kruskal-Wallis test (Table 1.2) demonstrate that flag condition, age, and sex significantly influence marathon performance. Age bin exhibited the largest effect, with a chi-square value of 7,436.804 (p \< 0.001), underscoring the critical role of age in shaping marathon performance. These findings highlight the natural decline in performance with increasing age, which is consistent with previous studies. The significant chi-square value for sex (836.146, p \< 0.001) reflects well-documented differences in endurance performance between men and women, with men typically achieving faster times across most age bins. Flag condition also played a significant role, with a chi-square value of 61.851 (p \< 0.001), indicating that environmental severity affects marathon outcomes.

The results of the pairwise Wilcoxon tests for age bins (Table 1.3) highlight significant differences in marathon performance between most age groups. Younger age bins (\<18, 18-25) consistently showed faster performances compared to older groups, with all comparisons yielding Bonferroni-adjusted p-values below 0.001. Performance differences were particularly pronounced when comparing middle-aged runners (36-45, 46-55) to older groups (56-64, 65+), reflecting the natural decline in endurance capabilities with age. Interestingly, the largest performance gaps were observed between the youngest (\<18) and oldest (65+) groups, underscoring the cumulative impact of aging on marathon times.

```{r}

# Kruskal-Wallis Tests
kw_flag <- kruskal.test(time_min ~ flag, data = cleaned_df)
kw_age_bin <- kruskal.test(time_min ~ age_bin, data = cleaned_df)
kw_sex <- kruskal.test(time_min ~ sex, data = cleaned_df)


# Pairwise Wilcoxon Tests
pw_flag <- pairwise.wilcox.test(
  cleaned_df$time_min,
  cleaned_df$flag,
  p.adjust.method = "bonferroni"
)

pw_age_bin <- pairwise.wilcox.test(
  cleaned_df$time_min,
  cleaned_df$age_bin,
  p.adjust.method = "bonferroni"
)

pw_sex <- wilcox.test(time_min ~ sex, data = cleaned_df)


# Format Kruskal-Wallis Results into a Table
kw_results <- data.frame(
  Variable = c("Flag", "Age Bin", "Sex"),
  Chi_Squared = c(kw_flag$statistic, kw_age_bin$statistic, kw_sex$statistic),
  p_value = c(kw_flag$p.value, kw_age_bin$p.value, kw_sex$p.value)
)

kw_table <- kw_results %>%
  gt() %>%
  tab_header(title = "Table 1.2: Kruskal-Wallis Test Results") %>%
  cols_label(
    Variable = "Variable",
    Chi_Squared = "Chi-Square",
    p_value = "p-value"
  ) %>%
  fmt_number(columns = c(Chi_Squared, p_value), decimals = 3)

kw_table

# Pairwise Wilcoxon Test for Age Bins
pw_age_bin <- pairwise.wilcox.test(
  cleaned_df$time_min,
  cleaned_df$age_bin,
  p.adjust.method = "bonferroni"
)

pw_age_bin_table <- as.data.frame(pw_age_bin$p.value) %>%
  rownames_to_column(var = "Comparison") %>%
  pivot_longer(cols = -Comparison, names_to = "Group", values_to = "Adjusted_p") %>%
  gt() %>%
  tab_header(
    title = "Table 1.3: Pairwise Wilcoxon Test Results for Age Bins",
    subtitle = "Bonferroni-Adjusted p-values"
  ) %>%
  cols_label(
    Comparison = "Age Bin Comparison",
    Group = "Group Compared",
    Adjusted_p = "Adjusted p-value"
  ) %>%
  fmt_number(columns = "Adjusted_p", decimals = 3)

pw_age_bin_table

```

Further analysis, as visualized in **Figure 1.2**, shows that peak performance age varies between men and women across different races. Men tend to reach their peak performance around 28-30 years old, and women tend to reach there peak around 28 to 36 years. One particularly interesting finding is that in Duluth, women’s best performance age extends to 36 years. This six-year gap compared to men, who generally peak around 29, raises questions about whether the summer conditions in Duluth might contribute to this outlier. This suggests that gender differences in thermoregulation may play a role in women reaching their peak later in Duluth races. Overall, the 26-35 age bin is the period of fastest marathon times for both men and women.

```{r fig.width=9, fig.height=6}

# average percent_cr for each age and sex
avg_percent_age <- cleaned_df %>%
  group_by(age, sex) %>%
  summarise(avg_percent_cr = mean(percent_cr, na.rm = TRUE), .groups = 'drop')

# calculate the average time each age, sex, and race
avg_time_age <- cleaned_df %>%
  group_by(age, sex, race) %>%
  summarise(avg_time = mean(time_min, na.rm = TRUE), .groups = 'drop')

# min percent for each age and sex
min_percent_by_sex <- avg_percent_age %>%
  group_by(sex) %>%
  summarise(min_percent_cr = min(avg_percent_cr),
            min_age = age[which.min(avg_percent_cr)],
            .groups = 'drop')


# minimum average time and corresponding age for each sex and race
min_time_by_sex_race <- avg_time_age %>%
  group_by(sex, race) %>%
  summarise(min_time = min(avg_time),
            min_age = age[which.min(avg_time)],
            .groups = 'drop')


# plot showing average time, showing minimum age by sex
age_time_avg <- ggplot(avg_time_age, aes(x = age, y = avg_time, color = sex)) +
  geom_point(shape = "square", size = .5) +
  geom_line(size = .25) + 
  geom_hline(data = min_time_by_sex_race, aes(yintercept = min_time, color = sex), linetype = "dashed") +  # Horizontal dashed line
  geom_text(data = min_time_by_sex_race, 
            aes(x = min_age, y = min_time, label = paste0(sex, " Min Age: ", min_age)),
            hjust = -1.5, vjust = 0, color = "black", size = 3.25, family = "serif") +  
  facet_wrap(~race) +  # Facet by race
  labs(x = "Age", y = "Average Marathon Time (minutes)", 
       title = "Figure 1.2: Average Marathon Performance by Age and Race",
       color = "Sex") +
  scale_color_manual(values = c("F" = "#BE398D", "M" = "#43A5BE")) +
  theme_classic(base_family = "serif") + 
  theme(legend.position = "bottom")


age_time_avg

```

In summary, these results support the hypothesis that aging impacts marathon performance differently for men and women. Both men and women experience declines in performance as they age. These findings offer valuable insights into the interplay between age and sex in long-distance running events.

```{r fig.width=10, fig.height=7}


```

## Aim 2: Effects of Environmental Conditions on Marathon Performance in Men and Women

For Aim 2, the goal was to determine how environmental factors influence marathon performance and whether these effects differ by age and gender. We determined 2 kinds of environmental conditions: weather and air quality. We first looked at the weather parameters.

### Relationship between Weather Variables and Marathon Performance

**Table 2.1** displays some summary statistics for all related weather variables.

```{r}
library(dplyr)
library(tidyr)
library(gt)
library(webshot2)

# Summarize the weather data by race
weather_summary_race <- cleaned_df %>%
  group_by(race) %>%
  summarize(
    `WBT (°C)` = paste0(round(median(wb_temp, na.rm = TRUE), 1), " (", 
                         round(quantile(wb_temp, 0.25, na.rm = TRUE), 1), ",", 
                         round(quantile(wb_temp, 0.75, na.rm = TRUE), 1), ")"),
    `DBT (°C)` = paste0(round(median(db_temp, na.rm = TRUE), 1), " (", 
                         round(quantile(db_temp, 0.25, na.rm = TRUE), 1), ",", 
                         round(quantile(db_temp, 0.75, na.rm = TRUE), 1), ")"),
    `RH (%)` = paste0(round(median(rel_humid, na.rm = TRUE), 1), " (", 
                      round(quantile(rel_humid, 0.25, na.rm = TRUE), 1), ",", 
                      round(quantile(rel_humid, 0.75, na.rm = TRUE), 1), ")"),
    `SR (W/m²)` = paste0(round(median(solar_rad, na.rm = TRUE), 1), " (", 
                         round(quantile(solar_rad, 0.25, na.rm = TRUE), 1), ",", 
                         round(quantile(solar_rad, 0.75, na.rm = TRUE), 1), ")"),
    `Wind (m/s)` = paste0(round(median(wind, na.rm = TRUE), 1), " (", 
                          round(quantile(wind, 0.25, na.rm = TRUE), 1), ",", 
                          round(quantile(wind, 0.75, na.rm = TRUE), 1), ")"),
    `WBGT (°C)` = paste0(round(median(WBGT, na.rm = TRUE), 1), " (", 
                          round(quantile(WBGT, 0.25, na.rm = TRUE), 1), ",", 
                          round(quantile(WBGT, 0.75, na.rm = TRUE), 1), ")")
  )

# Summarize the flag counts and percentages
flag_summary <- cleaned_df %>%
  group_by(race, flag) %>%
  summarize(flag_count = n(), .groups = "drop") %>%
  group_by(race) %>%
  mutate(flag_percent = round((flag_count / sum(flag_count)) * 100, 1)) %>%
  mutate(flag_summary = paste0(flag, ": ", flag_count, " (", flag_percent, "%)")) %>%
  summarize(flag_summary = paste(flag_summary, collapse = "; "), .groups = "drop")

# Merge weather summary with flag summary
merged_summary <- weather_summary_race %>%
  left_join(flag_summary, by = "race")

# Transpose the summarized data
transposed_summary <- merged_summary %>%
  pivot_longer(cols = -race, names_to = "Weather and Flag Variables", values_to = "Value") %>%
  pivot_wider(names_from = race, values_from = Value)

# Create a gt table from the transposed data
transposed_gt_table <- transposed_summary %>%
  gt() %>%
  tab_header(
    title = "Weather and Flag Summary by Race",
    subtitle = "Median (IQR) of Weather Variables and Flag Summary"
  ) %>%
  cols_label(
    `Weather and Flag Variables` = "Variables"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", align = "center"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_options(
    table.font.size = "small",
    heading.title.font.size = 16,
    heading.subtitle.font.size = 12
  ) %>%
  tab_footnote(
    footnote = "WBT: Wet Bulb Temp, DBT: Dry Bulb Temp, RH: Relative Humidity, 
    SR: Solar Radiation, WBGT: Wet Bulb Globe Temp. Flag Summary includes counts and percentages of each flag type.",
    locations = cells_column_labels(columns = everything())
  ) %>%
  tab_style(
    style = cell_text(size = px(10)), 
    locations = cells_footnotes()
  )

# View the transposed gt table
transposed_gt_table

# Save the gt table as a PNG
gtsave(
  data = transposed_gt_table,
  filename = "transposed_weather_and_flag_summary.png"
)
```

From this table, we can see some differences in weather conditions across different races. For example, Duluth (Grandma's), held during the summer, consistently experiences the highest median wet bulb temperature (14.3°C) and dry bulb temperature (18.1°C), along with the highest wet bulb globe temperature (WBGT) at 18.1°C. This race also has the highest relative humidity (60.3%), which, combined with high solar radiation and temperatures, suggests more challenging conditions for runners compared to other races. By contrast, New York and Boston races tend to have windier conditions, with median wind speeds of 11 and 12 m/s, respectively, compared to other races. These factors highlight Duluth’s more intense environmental conditions, which may contribute to longer marathon times due to heat and humidity. We further explored the relationship between weather and marathon performance in **Figure 2.1**, where we incorporated flag conditions as a categorical representation of WBGT. Flag conditions serve as indicators of temperature-related risk, with white flags representing the coldest conditions, followed by green for moderate conditions, yellow for warmer, and red for the hottest and most hazardous conditions. The differences in weather conditions across races, particularly Duluth's harsher environment, may impact runners' performance and influence the distribution of flag conditions.

```{r fig.width=10, fig.height=7}
cleaned_df <- cleaned_df %>%
  mutate(flag = factor(flag, levels = c("White", "Green", "Yellow", "Red")))

# Create the boxplot with ordered legend
boxplot_age_bin <- ggplot(cleaned_df, aes(x = age_bin, y = percent_cr, fill = flag)) +
  geom_boxplot(outlier.size = 0.5, size = 0.25) +
  facet_wrap(~race, ncol = 2) +  # Facet by race
  scale_fill_manual(values = flag_colors) +
  labs(title = "Figure 2.1: Marathon Performance by Age Bin and Flag Condition",
       x = "Age Bin",
       y = "Marathon Time (% Course Record)",
       fill = "Flag Condition") +
  theme_classic(base_family = "serif") +
  theme(
    panel.grid.major = element_line(size = 0.25),
    legend.position = "bottom"
  )

boxplot_age_bin


```

These results show that marathon times increase with age across all races, especially in the older age groups (65+). The red flags, indicating the hottest conditions, are generally associated with the highest marathon times, particularly in the 65+ age group, where there is considerable variation. The white and green flags, which represent cooler conditions, tend to have the fastest times across most age bins, highlighting the impact of favorable weather conditions on performance. We observe that New York does not have any races with red flags from the given years in the data, and Duluth does not have any white flag races. The latter is true, most likely because it takes place in the summer, where white flag conditions are extremely rare in that location. Also, we can see missingness in the \< 18 age group in New York and Boston, where one could infer that they are missing because one must be 18+ to qualify for those marathons, or perhaps there is some other reason we don;t observe people from that age range in our data. **Figure 2.2** further illustrates the differences in flag distributions across races.

### Relationship Between Air Quality and Marathon Performance

The second type of environmental factor that we explored was air quality, measured by ozone in parts per million (PPM). In **Figure 2.3**, we visualized the marathon times in percent course record across different PPM levels that were observed in the data, stratified by age range. For all age groups, we mostly saw a flat regression line fit to the data, indicating a weak linear relationship. There were minor fluctuations in older groups but they were still relatively flat and uninformative. From this visual, we assume that air quality most likely does not have a significant effect on marathon times for neither men nor women. Because of this, we will focus on discovering which weather parameters impact marathon performance the most in aim 3.

```{r fig.width=10, fig.height=7}

ppm_percent_cr <- ggplot(cleaned_df, aes(x = avg_ppm, y = percent_cr, color = sex)) +
  geom_point(alpha = 0.1) + 
  geom_smooth(se = FALSE, size = 1) +
  facet_wrap(~age_bin, scales = "free_y", nrow = 4) +
  labs(title = "Figure 2.3: Air Quality (Ozone PPM) vs Marathon Time by Sex & Age Group",
       x = "Ozone PPM",
       y = "Marathon Time (% Percent CR)",
       color = "Sex") +
  scale_color_manual(values = c("F" = "#BE398D", "M" = "#43A5BE")) +
  theme_classic(base_family = "serif") +
  theme(legend.position = "bottom")

ppm_percent_cr

```

## Aim 3

For Aim 3, the analysis shifted from comparing the effects of environmental conditions across demographic groups to identifying which specific weather variables had the most significant impact on overall marathon performance. As shown in the correlation heatmap (**Figure 3.1**), WBGT and dry bulb temperature were the most impactful environmental variables, but their correlations with marathon times were modest (around 0.10 for WBGT and DBT in some races). Given the relatively low correlations with other variables like solar radiation and dew point, WBGT emerged as the primary continuous measure of environmental stress, while flag conditions were used as a categorical measure for weather severity.

```{r}
cor_vars <- cleaned_df %>%
  select(time_min, dew_point, rel_humid, globe_temp, wb_temp, db_temp, 
         solar_rad, wind, WBGT)

cor_matrix <- cor(cor_vars, use = "complete.obs")
melted_cor_matrix <- melt(cor_matrix)

cor_plot <- ggplot(data = melted_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#090C9B", high = "#BE398D", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  labs(title = "Figure 3.1: Correlation Between Weather Variables and Performance") +
  theme_minimal(base_family = "serif") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), legend.position = "bottom")

cor_plot
```

The bar plot showing correlations by race (**Figure 3.2**) highlighted the variation in these relationships across different marathons. For instance, Boston exhibited a 0.10 correlation between marathon times and DBT, while Duluth and Chicago had slightly higher correlations between WBGT and performance. These relatively small correlation values suggested that no single environmental variable had a dominant effect on marathon times, but WBGT consistently emerged as the most relevant continuous measure.

```{r}

race_colors = c("Boston" = "#005A36", "New York" = "#090C9B", "Chicago" = "#BE398D",
                "Duluth (Grandma's)" = "#43A5BE", "Twin Cities" = "#C1B8FF")

# group by race, calculate correlations between weather and marathon time
correlations_by_race <- cleaned_df %>%
  group_by(race) %>%
  summarize(
    wb_temp_corr = cor(time_min, wb_temp, use = "complete.obs"),
    db_temp_corr = cor(time_min, db_temp, use = "complete.obs"),
    rel_humid_corr = cor(time_min, rel_humid, use = "complete.obs"),
    solar_rad_corr = cor(time_min, solar_rad, use = "complete.obs"),
    wind_corr = cor(time_min, wind, use = "complete.obs"),
    WBGT_corr = cor(time_min, WBGT, use = "complete.obs")
    
  ) %>%
  gather(key = "weather_var", value = "correlation", -race)

# plot the correlations by race
corr_race_plot <- ggplot(correlations_by_race, aes(x = weather_var, y = correlation, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  scale_fill_manual(values = race_colors) +
  labs(
    title = "Figure 3.2: Correlations Between Weather Variables and Marathon Times",
    x = "Weather Variable",
    y = "Correlation with Marathon Time"
  ) +
  theme_classic(base_family = "serif") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")


corr_race_plot
# Function to calculate Fisher z-transformed confidence intervals
cor_with_ci <- function(x, y) {
  n <- sum(!is.na(x) & !is.na(y))  # Number of complete observations
  r <- cor(x, y, use = "complete.obs")  # Correlation coefficient
  fisher_z <- atanh(r)  # Fisher z-transformation
  se <- 1 / sqrt(n - 3)  # Standard error
  ci_lower <- tanh(fisher_z - 1.96 * se)  # Lower bound
  ci_upper <- tanh(fisher_z + 1.96 * se)  # Upper bound
  data.frame(correlation = r, ci_lower = ci_lower, ci_upper = ci_upper)

}

# Calculate correlations and confidence intervals by race and weather variable
correlations_with_ci <- cleaned_df %>%
  group_by(race) %>%
  summarize(
    wb_temp = cor_with_ci(time_min, wb_temp),
    db_temp = cor_with_ci(time_min, db_temp),
    rel_humid = cor_with_ci(time_min, rel_humid),
    solar_rad = cor_with_ci(time_min, solar_rad),
    wind = cor_with_ci(time_min, wind),
    WBGT = cor_with_ci(time_min, WBGT)
  ) %>%
  pivot_longer(cols = -race, names_to = "weather_var", values_to = "correlation_data") %>%
  unnest(correlation_data)

corr_race_plot <- ggplot(correlations_with_ci, aes(x = weather_var, y = correlation, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper), 
    position = position_dodge(width = 0.9), 
    width = 0.2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  scale_fill_manual(values = race_colors) +
  labs(
    title = "Figure 3.2: Correlations Between Weather Variables and Marathon Times",
    x = "Weather Variable",
    y = "Correlation with Marathon Time"
  ) +
  theme_classic(base_family = "serif") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")

corr_race_plot

```

The pairwise Wilcoxon test results for flag conditions, with Bonferroni-adjusted p-values, reveal significant differences in marathon performance among several flag comparisons. Notably, the Red flag condition showed highly significant differences compared to all other flags, including White (p \< 0.0001), Green (p \< 0.0001), and Yellow (p = 0.0182). This indicates that the Red flag condition, representing the most severe weather conditions, is associated with the largest deviations in marathon performance. The Yellow flag also displayed significant differences when compared to the White flag (p = 0.0007), highlighting the impact of moderate weather conditions on performance. However, no significant differences were observed between Green and White flags (p = 0.1078) or between Yellow and Green flags (p = 0.1972), suggesting that performance differences under these conditions are less pronounced. These findings suggest that as weather conditions worsen, particularly under Red and Yellow flags, marathon performance is increasingly impacted, likely due to adverse environmental factors such as heat and humidity.

```{r}

# Perform Kruskal-Wallis test for flag conditions
kw_test_flag <- kruskal.test(percent_cr ~ flag, data = cleaned_df)


# Pairwise Wilcoxon test with Bonferroni correction
wilcox_results_flag <- pairwise.wilcox.test(
  cleaned_df$percent_cr,
  cleaned_df$flag,
  p.adjust.method = "bonferroni"
)

# Convert Wilcoxon test results to a tidy format
wilcox_results_flag_tidy <- as.data.frame(as.table(wilcox_results_flag$p.value)) %>%
  filter(!is.na(Freq)) %>% # Remove NA values
  rename(Comparison = Var1, Group_Compared = Var2, Adjusted_P = Freq) %>%
  mutate(Adjusted_P = round(Adjusted_P, 4)) # Round p-values


wilcox_table_flag <- wilcox_results_flag_tidy %>%
  gt() %>%
  tab_header(
    title = "Pairwise Wilcoxon Comparisons for Flag Conditions",
    subtitle = "Bonferroni-Adjusted P-values"
  ) %>%
  cols_label(
    Comparison = "Comparison",
    Group_Compared = "Group Compared",
    Adjusted_P = "Adjusted P-value"
  ) %>%
  fmt_number(
    columns = Adjusted_P,
    decimals = 4
  ) %>%
  tab_options(
    table.font.size = "medium",
    heading.title.font.size = 16
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  )


wilcox_table_flag

```

When evaluating weather impact across age bins, older runners showed the greatest sensitivity to adverse weather. For instance, in the 65+ group, marathon times were significantly slower under red and yellow flag conditions, while younger age groups (such as 18-25) showed less variation in their times across flag categories (**Figure 2.1**). This difference highlights that flag conditions, particularly red and yellow flags, have a much more pronounced effect on older runners than on their younger counterparts.

In summary, Aim 3 demonstrated that WBGT and flag conditions had the most significant impacts on marathon performance. While the correlations were relatively modest, the substantial increases in marathon times under more severe flag conditions, particularly for older runners, confirmed the importance of WBGT as the key continuous variable and flag conditions as the primary categorical measure of weather impact.

# Discussion and Conclusion

The exploratory analysis confirm well-established trends, such as the decline in performance with age, but also reveal nuances, particularly how men and women are differently affected by aging and environmental stressors. While both men and women show peak marathon performance in their late 20s to early 30s, the decline in performance with age is more pronounced in women, particularly in the 65+ age group. The growing gender gap in older runners suggests that physiological factors like muscle loss and thermoregulation may affect women more severely as they age. Notably, Duluth’s outlier, where women seem to peak six years later than men, raises questions about whether this is linked to the race's summer heat, which might affect women’s performance differently.

The environmental conditions, particularly WBGT and flag conditions, played a significant but relatively modest role in affecting performance. Duluth’s high WBGT, for example, explains some of the slower marathon times compared to races held in cooler seasons, like Boston or New York. The results also showed that older runners were more sensitive to adverse weather.

Interestingly, our analysis showed no significant relationship between ozone levels and marathon performance. This could be due to generally lower ozone levels during the events or the brief exposure time for runners.

Future research could look into the physiological mechanisms driving performance declines, particularly the gendered responses to heat and age. As climate change intensifies, it will also be critical to examine how rising temperatures and increasingly erratic weather conditions may further impact marathon performance and the safety of runners.

# **Limitations**

The methodological approach used in this study, while appropriate for the dataset and research objectives, has inherent limitations. First, the reliance on non-parametric methods, such as the Kruskal-Wallis and pairwise Wilcoxon tests, limits the ability to adjust for potential confounding variables or interactions between demographic and environmental factors. While these methods are robust against violations of normality and homogeneity of variances, they do not allow for the inclusion of covariates or the modeling of complex relationships that may influence marathon performance.

Future research should consider employing more advanced statistical methods, such as multivariable regression or machine learning models, to account for potential confounders and interactions. ,

# References:

1.  Ely, B. R., Cheuvront, S. N., Kenefick, R. W., & Sawka, M. N. (2010). Aerobic performance is degraded, despite modest hyperthermia, in hot environments. Med Sci Sports Exerc, 42(1), 135-41.

2.  Ely, M. R., Cheuvront, S. N., Roberts, W. O., & Montain, S. J. (2007). Impact of weather on marathon-running performance. Medicine and science in sports and exercise, 39(3), 487-493.

3.  Kenney, W. L., & Munce, T. A. (2003). Invited review: aging and human temperature regulation. Journal of applied physiology, 95(6), 2598-2603.

4.  Besson, T., Macchi, R., Rossi, J., Morio, C. Y., Kunimasa, Y., Nicol, C., ... & Millet, G. Y. (2022). Sex differences in endurance running. Sports medicine, 52(6), 1235-1257.

5.  Yanovich, R., Ketko, I., & Charkoudian, N. (2020). Sex differences in human thermoregulation: relevance for 2020 and beyond. Physiology, 35(3), 177-184.

\newpage

# Code Amoppendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
